If you see something, then say something to others: Complex contagions and the socially-contingent correction of misinformation

### 2018-02-14 manuscript_1.0.tex
probably the last update on section 3

% Biblatex
\usepackage[style=apa,sortcites=true,sorting=nyt,backend=biber,uniquename=false]{biblatex}
% make in-text citations clickable
\usepackage[colorlinks=true]{hyperref}
% this removes annoying colors for in-text citation entries 
\usepackage{xcolor}
\hypersetup{  
    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}

@@

\title{If you see something, then say something to others: A simple contagion of misinformation and a complex contagion of socially-contingent corrections}

@@

\note{Word count: 3284 words %\quickwordcount \\
	  Draft date: \today \\
      \addvspace{.25in} 
      \textbf{Draft in progress. Please do not cite without permission.} \\
      Please direct any questions and inquiries to \href{mailto:hyunjin.song@univie.ac.at}{hyunjin.song@univie.ac.at}}

@@

	An effective deliberation among public is regarded as a keystone of thriving democracies, and modern political systems squarely depend  on  informed  decisions of citizens in that regard \parencite{carpini1996}. Yet, a propagation of rumors, misinformation, and so-called \enquote{fake news} on those platforms becomes an increasing concern for the public and policy makers alike \parencite{allcott2017social, lazer2017combating}, %There are abundant, yet still fragmentary, evidence of spreads of unsubstantiated, often factually dubious (mis)information through digital social networks that potentially affecting millions of citizens across the globe -- 
    as evidenced in recent 2016 U.S. presidential election \parencite{guess2018selective, giglietto2016fakes, allcott2017social} and in Brexit votes \parencite{nyt_2017}. While a wide circulation of factually dubious information is not entirely new to political arena, a growing trend of digitally disseminated rumors and misinformations -- often termed as a \enquote{fake news} phenomenon -- is increasingly recognized as a serious threat to liberal democratic societies \parencite{allcott2017social, lazer2017combating}. Either based on unsubstantiated rumors or based on factually wrong beliefs, many of the misinformed behave differently than those who are accurately informed \parencite{kuklinski2000misinformation}. They often disagree about basic facts about many political and public issues \parencite[e.g.,][]{nisbet2015partisan}, and continue to believe and rely on such false information when making political judgments \parencite{nyhan2010corrections,thorson_2016}.

@@

\section{A Psychology of Fake News, Misperceptions, and Corrections}

@@

    Moreover, due to inherent partisan motivated directionality, the threshold of actually \emph{adapting} such a false yet pro-attitudinal claim (i.e., believing attitudinally congruent misinformation) may also exhibit similar low-threshold properties, although the actual threshold for believing misinformation would be bit higher compared to that of mere exposure and awareness \parencite[e.g.,][]{Monsted_plos2017}, and also subject to some individual differences \parencite[e.g.,][]{weeks2015emotions}. As previously suggested, people tend to directionally process political information \parencite{taber2006}, and they tend to easily remember and recall attitudinally-consistent information than uncongenial ones. Therefore, at least for citizens who find a given misinformation to be congenial to their prior partisan identities, \emph{adapting} such false yet pro-attitudinal claims also does not require high number of repeated exposure nor independent, multiple exposure to different sources supporting such claims. Indeed, many of the prior empirical studies support this perspective \parencite[e.g.,][]{garrett2016driving, kuklinski2000misinformation, nisbet2015partisan, flynn2017nature}.

@@

	There are at least several other reasons why socially-based correction messages, especially from one's peers, might be more effective than a single-shot, \emph{asocial} correction message from more distanced sources (such as fact-checking organizations or mere strangers). First, people may evaluate information coming from their peers to be more credible and trustworthy \parencite[e.g.,][]{metzger2010social}, and more willing to deliberate with their close social contacts \parencite{morey2012matters}. Research suggests that while individuals rather maintain much flexible attitudes as long as their social affiliation goals are met \parencite{levitanspy2017}, reputational risks and social accountability of rejecting corrective messages run high for more close social contacts compared to more distant sources such as strangers \parencite{margolin2017}.

@@

\section{Analytical Challenges in Studying (Mis)information Diffusion Within Social Networks}

	If the diffusion of (mis)information and adoption of corrective messages may non-trivially dependent upon a structure of a given network and its attitudinal composition, then how a typical (online) social network is structured in terms of its topological features, and how pervasive is \enquote{attitudinal} homophily on such a social network? How such structural features affect the overall diffusion dynamics empirically? This is indeed an important question to ask, since the flow of information (either misinformation or its correction) and its adoption are ultimately structured by how individuals are connected with each other in a given network.

    A frequent and recurring theme for (attitudinal) makeup of citizens' social networks and its consequences, among scholars and general publics alike, is that most of the citizens today are put into a \enquote{echo chamber} or a \enquote{filter bubble} that insulate themselves from competing viewpoints and attitude-discrepant information \parencite[e.g.,][]{lewandowsky2017beyond, Bessi_2015, del2016spreading}. Yet, as \citeauthor{Garrett2017distraction} (\citeyear{Garrett2017distraction}) puts it, \enquote{there is ample evidence that \emph{exposure} [emphasis added] echo chambers are not a typical part of Internet users' experience} (p. 370). Most of citizens are appear to be embedded in sufficiently diverse social networks, either online \parencite[e.g.,][]{Bakshy1130, messing2014selective} or offline \parencite{song2015uncovering, lazer2010coevolution}, documenting a substantial level of exposure to political difference in citizen’s everyday political interactions. This suggests that popular claims of echo chambers or filter bubbles are less warranted than it is often assumed.\footnote{		Indeed, in a completely segregated network where cross-ideological links are not present, a simple diffusion of partisan misinformation is not likely to saturate the entire network, since at least some segments of populations are never exposed to such information due to the lack of cross-ideological exposure. In light of our discussion, an \enquote{exposure} echo chamber actually \emph{prohibits} global-scale (mis)information propagation.} However, one should also bear in mind that evidence of exposure to a counter-attitudinal messages does not provide enough evidence for how such messages are actually interpreted and adopted \parencite[e.g.,][]{nyhan2017taking, Garrett2017distraction}. Indeed, many of the prior studies on this topic relies on observational evidence regarding message \emph{exposure patterns} alone, either based on participants' self-reports or based on messages posted (sometimes along with engagement data such as \enquote{likes} or \enquote{shares}) in online social networks \parencite[e.g.,][]{Bakshy1130}, not based on how an individuals actually process and interpret such information. As suggested by elsewhere, attitudinal composition of one's network may fundamentally alter how (identical) messages are interpreted and received by an individual \parencite{Garrett2017distraction, levitan2008resistance}. As experimental evidence to date suggests, the mere presence of ideologically diverse \emph{exposure} does not automatically translate into the possibility that such exposure would promote more balanced judgments. In contrast, experimental evidence provide more detailed picture of how such (counter-attitudinal) messages are actually processed and adopted \parencite[e.g., ][]{messing2014selective, nyhan2017taking, Wood2018}. Yet as previously suggested, a typical experimental approach tends to be misspecified in terms of important social dynamics surrounding the simple vs. complex contagion of misinformation and (counter-attitudinal) corrective messages \parencite{margolin2017, Centola2010Sience}. In addition, designing a realistic experiment involving real-world social interactions often involves significant practical \parencite[e.g.,][]{bond_61million} and ethical challenges \parencite[e.g.,][]{Kramer8788} for researchers.

    Similarly to attitudinal composition of citizen's social networks, the question of how exactly the structure of citizens' interaction exhibit certain topological properties have attracted considerable interests among scholars. Prior observations on this topic tend to suggests that large-scale online networks exhibit \enquote{small-world}-like properties \parencite{kumar2010structure, ugander2011anatomy}. A small world is characterized by high connectivity and relatively short average distance among most nodes -- most nodes can be reached from every other node by a small number of hops or steps -- compared to pure random graphs. This is due to the fact that a small minority of \enquote{hub} nodes concentrates a disproportionate number of links, providing a global bridge between smaller, more strongly interconnected local clusters \parencite{barabasi2004linked}. This structure is known for expediting simple information to be spread globally \parencite{granovetter1977strength, Bakshy_2012}, yet it also tends to facilitate complex contagion since the existence of tightly knit neighborhoods supports complex-type behavioral propagation that is difficult or considered to be run counter to prevalent norms and values \parencite{Centola2010Sience}.

    %\enquote{Despite the scientific and practical importance...an empirical test of these predictions has not been possible, because it requires the ability to independently vary the topological structure of a social network}

%\subsection{}    
%    \enquote{While much of the literature on networks treats the topological features of social structure as given quantities, networks do not emerge \emph{ex nihilo}, but are endogenously formed}

>>>>

% Biblatex
\usepackage[style=apa,sortcites=true,sorting=nyt,backend=biber,uniquename=false]{biblatex}
\AtEveryBibitem{%
  \clearfield{issn} % Remove issn
  \clearfield{doi} % Remove doi
  }
% make in-text citations clickable
\usepackage[colorlinks=true]{hyperref}
% this removes annoying colors for in-text citation entries 
\usepackage{xcolor}
\hypersetup{  
    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}

@@

\title{If you see something, then say something to others: Complex contagions and the socially-contingent correction of misinformation}

@@

\note{Word count: 3599 words \\ %\quickwordcount \\
    Draft date: \today \\
      \addvspace{.25in} 
      \textbf{Draft in progress. Please do not cite without permission.} \\
      Please direct any questions and inquiries to \href{mailto:hyunjin.song@univie.ac.at}{hyunjin.song@univie.ac.at}}

@@

  An effective deliberation among public is regarded as a keystone of thriving democracies, and modern political systems squarely depend  on  informed  decisions of citizens in that regard \parencite{carpini1996}. Yet, a propagation of rumors, misinformation, and so-called \enquote{fake news} on those platforms becomes an increasing concern for the public and policy makers alike \parencite{allcott2017social, lazer2017combating}, as evidenced in recent 2016 U.S. presidential election \parencite{guess2018selective, giglietto2016fakes, allcott2017social} and in Brexit votes \parencite{nyt_2017}. While a wide circulation of factually dubious information is not entirely new to political arena, a growing trend of digitally disseminated rumors and misinformations -- often termed as a \enquote{fake news} phenomenon -- is increasingly recognized as a serious threat to liberal democratic societies \parencite{allcott2017social, lazer2017combating}. Either based on unsubstantiated rumors or based on factually wrong beliefs, many of the misinformed behave differently than those who are accurately informed \parencite{kuklinski2000misinformation}. They often disagree about basic facts about many public issues, and continue to believe and rely on such false information when making political judgments \parencite{nyhan2010corrections,thorson_2016}.

@@

\section{Psychological Underpinnings of Fake News, Misperceptions, and Corrections}

@@

    Moreover, due to inherent partisan motivated directionality, the threshold of actually \emph{adapting} such a false yet pro-attitudinal claim (i.e., believing attitudinally congruent misinformation) may also exhibit similar low-threshold properties, although the actual threshold for believing misinformation would be bit higher compared to that of mere exposure and awareness \parencite[e.g.,][]{Monsted_plos2017}, and also subject to some individual differences \parencite[e.g.,][]{weeks2015emotions}. As previously suggested, people tend to directionally process political information \parencite{taber2006}, and they tend to easily remember and recall attitudinally-consistent information than uncongenial ones. Therefore, at least for citizens who find a given misinformation to be congenial to their prior partisan identities, \emph{adapting} such false yet pro-attitudinal claims also does not require high number of repeated exposure nor independent, multiple exposure to different sources supporting such claims. Indeed, many of the prior empirical studies support this perspective \parencite[e.g.,][]{garrett2016driving, kuklinski2000misinformation, flynn2017nature}.

@@

  There are also at least several other reasons why socially-based correction messages, especially from one's peers, might be more effective than a single-shot, \emph{asocial} correction message from more distanced sources (such as fact-checking organizations or mere strangers). First, people may evaluate information coming from their peers to be more credible and trustworthy \parencite[e.g.,][]{metzger2010social}, and more willing to deliberate with their close social contacts \parencite{morey2012matters}. Research suggests that while individuals rather maintain much flexible attitudes as long as their social affiliation goals are met \parencite{levitanspy2017}, reputational risks and social accountability of rejecting corrective messages run high for more close social contacts compared to more distant sources such as strangers \parencite{margolin2017}.

@@

\section{Observational Challenges in Studying Misperception Within Social Networks}

  If the diffusion of (mis)information and adoption of corrective messages may non-trivially dependent upon a structure of a given network and its attitudinal composition, then how a typical (online) social network is structured in terms of its topological features, and how pervasive is \enquote{attitudinal} homophily on such a social network? How such structural features affect the overall diffusion dynamics empirically? This is indeed important questions to ask, since the flow of information (either misinformation or its correction) and its adoption are ultimately structured by how individuals are connected with each other in a given network. However, it is surprisingly difficult to establish convincing evidence of the impact of structural properties of a given network and its attitudinal composition in complex contagion dynamics using purely observational and experimental approaches.

    A frequent and recurring theme for (attitudinal) a makeup of citizens' social networks and its consequences, among scholars and general publics alike, is that most of the citizens today are put into a \enquote{echo chamber} or a \enquote{filter bubble} that insulate themselves from competing viewpoints and attitude-discrepant information \parencite[e.g.,][]{lewandowsky2017beyond, del2016spreading}. Yet, as \citeauthor{Garrett2017distraction} (\citeyear{Garrett2017distraction}) puts it, \enquote{there is ample evidence that \emph{exposure} [emphasis added] echo chambers are not a typical part of Internet users' experience} (p. 370). Most of citizens are appear to be embedded in sufficiently diverse social networks, showing a substantial level of exposure to political difference online \parencite[e.g.,][]{Bakshy1130, messing2014selective} and offline \parencite[e.g.,][]{huckfeldt2004disagreement}. More importantly, those studies do not find compelling evidence that ordinary citizens' network constructions are primarily driven by purposive political homophily that limit their exposure to only congenial political perspectives \parencite{song2015uncovering, lazer2010coevolution}. This suggests that popular claims of echo chambers or filter bubbles are often overstated.\footnote{    Indeed, in a completely segregated network where cross-ideological links are not present, a simple diffusion of partisan misinformation is not likely to saturate the entire network, since at least some segments of populations are never exposed to such information due to the lack of cross-ideological exposure. In light of our discussion, an \enquote{exposure} echo chamber actually \emph{prohibits} global-scale (mis)information propagation.}

    However, one should also bear in mind that the evidence concerning mere \emph{exposure} to counter-attitudinal messages does not provide enough indications of how such messages are actually cognitively processed and interpreted \parencite[e.g.,][]{nyhan2017taking, Garrett2017distraction}. What we know about citizens' exposure to disagreement in their social networks typically has relied on observational evidence, either based on participants' self-reports regarding their patterns of social interactions \parencite[e.g.,][]{song2015uncovering, lazer2010coevolution} or based on messages posted (sometimes along with engagement data such as \enquote{likes} or \enquote{shares}) in online social networks \parencite[e.g.,][]{Bakshy1130}. While such evidence \emph{do} suggests that citizens are indeed exposed to competing viewpoints in their daily lives, they are typically rather silent about how an individuals actually process and interpret such information at the time of exposure.\footnote{  Recent studies try to overcome problems of typical observational studies by relying on time-stamped, detailed digital footprint data \parencite[e.g.,][]{guess2018selective, margolin2017}, yet such studies also suffer from similar weakness, in that they cannot provide strong causal evidence concerning individuals' actual cognitive processing of counter-attitudinal messages beyond what is available in observational data.} As experimental evidence to date suggests, the mere presence of ideologically diverse \emph{exposure} does not automatically translate into the possibility of more balanced judgments. Experimental evidence in this regard can provide more detailed pictures of how such (counter-attitudinal) messages are actually selected, processed, and adopted by an individual given exposure settings \parencite[e.g., ][]{messing2014selective, nyhan2017taking, Wood2018}. Yet as previously suggested, a typical experimental approach tends to be misspecified in terms of crucial social dynamics surrounding simple vs. complex contagion of misinformation and correction messages within a naturally-occurring social network \parencite{margolin2017, Centola2010Sience}. Moreover, designing a realistic experiment involving real-world social interactions often involves significant practical \parencite[e.g.,][]{bond_61million} and ethical challenges \parencite[e.g.,][]{Kramer8788} for researchers.

    Similarly to attitudinal composition of citizen's social networks, the question of how exactly the structure of citizens' social networks exhibits certain topological properties have attracted considerable interests among scholars. Prior observations on this topic suggests that large-scale (online) networks tend to exhibit \enquote{small-world} like properties \parencite{kumar2010structure, ugander2011anatomy}. A small world is characterized by its typical structure where most nodes can be reached from every other nodes by a small number of steps compared to a pure random graph of the same size. This is due to the fact that a small minority of \enquote{hub} nodes posses a disproportionate number of links, providing a global bridge between smaller, more strongly interconnected local clusters \parencite{barabasi2004linked}.

    This structure has traditionally been known for information to be spread globally through such hubs \parencite[e.g.,][]{Bakshy_2012}, following \citeauthor{granovetter1977strength}'s (\citeyear{granovetter1977strength}) seminal account of strength of weak ties. However, emerging evidence suggests that such a small-world structure may hinder a given diffusion process to globally saturate the network, since such hubs can serve as a \enquote{bottleneck} of a contagion process, therefore diffusions are stay localized and compartmentalized \parencite[e.g.,][]{banos2013diffusion, centola2007complex}. Similarly, while the existence of tightly-knit neighborhoods in a typical small-world network supports a complex-type behavioral propagation \parencite[i.e., an innovation that runs counter to prevalent norms and values:][]{Centola2010Sience}, the existence of hubs may not favor complex contagion to be globally propagated into a given network. This is due to the fact that such hubs require a much higher number of active neighbors to adopt a controversial innovation (i.e., complex contagions), and even if such hubs are activated, they alone cannot further activate their immediate neighbors to adopt such controversial innovations in the absence of other local nodes that provide multiple independent reinforcements \parencite{Centola2007449}. Therefore, whether topological properties of a given network actually hinder or facilitate a contagion process may critically dependent upon specific thresholds values for complex contagions \parencite[e.g.,][]{Centola2007449}, which is often hard to be tractable empirically.

    Moreover, especially for the case of misinformation and its corrections, a simple contagion (of misinformation) and a complex contagion (of correction messages) may simultaneously affect a local-level attitudinal composition of one's neighbors \parencite[i.e., those who endorse and believe misinformation vs. those who do not: e.g.,][]{campbell2013complex}. Last but not least, all of aforementioned factors are endogenously determined through prior states of networks and their topological properties over time. As a consequence, a proper identification regarding the exact impacts of its attitudinal composition and structural properties of an given network often necessitates systematic comparisons between the observed network and plausible counterfactuals. Doing so further requires (a) the ability to compared different thresholds values for complex contagions, and (b) the ability to independently manipulate topological structures of a social network \parencite[e.g.,][]{Centola2010Sience, Centola2007449}. For this reason, a rigorous empirical test of such factors is often practically impossible for observational studies despite its scientific and practical importance.
